Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-11-12 analysis/paper/paper.Rmd
edits to Discussion; added 5 packages to load-libraries

```{r load-libraries}
library(tidyverse)
library(magrittr)
library(janitor)

@@

The Upper Salmon River major population group (MPG) supports eight extant spring/summer Chinook Salmon *Oncorhynchus tshawytscha* populations including Salmon River (above Redfish Lake), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. At least five of these eight populations must meet criteria set forth by @McElhany2000 and @ICTRT2007 for the MPG to be considered viable and for the recovery of the Snake River Evolutionary Significant Unit (ESU). Populations within the ESU have substantial cultural value, support downriver mainstem Snake and Columbia River commercial and subsistence fisheries, and support local fisheries and economies in years with sufficient abundance. All populations within the Upper Salmon River MPG have become depleted in recent decades. Declines in survival of juvenile Chinook Salmon have been attributed to the removal of beavers from the landscape (fur trade), mining activities, river simplification, water withdrawals, logging activities, urbanization, avian predation, proliferation of non-native species (e.g., non-native coastal rainbow trout *O. mykiss irideus* and brook trout *Salvelinus fontinalis*), warming streams and rivers, and modifications to downriver migration corridors (e.g., from hydropower projects). Moreover, the abundance of returning adults are further impacted by ocean and downriver harvests, poor ocean conditions, and changes to the spawning migration corridor. Each of these factors have contributed, to varying and unknown extents, to reduced adult escapement, the primary metric used to assess population viability. In response to the decline in Chinook Salmon abundance from the myriad human activities and associated habitat degradation, action agencies have attempted to improve juvenile survival and adult spawning conditions by investing in the rehabilitation of tributary ecosystems.

@@

```{r schnabel-estimator}
# using Schnabel model
N_mods = cmr_df %>%
  filter(species == "Northern Pikeminnow") %>%
  # filter out spring 2021 (no recapture)
  filter(event_name != "Spring_2021") %>%
  group_by(event_name, species, date, fish_status, capture_type) %>%
  summarise(n_fish = n(),
            .groups = "drop") %>%
  group_by(event_name, species, date) %>%
  summarise(n = sum(n_fish),
            m = sum(n_fish[capture_type == "Recapture"]),
            R = sum(n_fish[fish_status == "Live"]),
            dead_fish = sum(n_fish[fish_status == "Dead"]),
            .groups = "drop") %>%
  group_by(event_name, species) %>%
  nest() %>%
  mutate(mod = map(data,
                   .f = function(x) {
                     with(x,
                          mrClosed(n = n,
                                   m = m,
                                   R = R,
                                   method = "Schnabel",
                                   chapman.mod = TRUE))
                   }),
         N = map_dbl(mod,
                     .f = summary),
         CI = map(mod,
                  .f = confint))

@@

We estimated the population size of northern pikeminnow in the Deadwater Slough and during the fall emigration period to be greater than ***xx,xxx***. And although we did not attempt to estimate population size during the spring emigration period, the relative abundance measured as CPUE was comparable to that during the fall sampling periods (update statement later). That fall population size estimate translates to ***xxx*** northern pikeminnow per 100 m or ***xxx*** pikeminnow per 100 m^2^. For context, that is similar/more/less than estimates from the Columbia River where predation impacts have been demonstrated as substantial and bounty programs have been implemented to reduce northern pikeminnow abundance. Given that estimate of abundance and some assumptions, we estimated that northern pikeminnow potentially consume between ***xx,xxx*** and ***xx,xxx*** juvenile Chinook salmon during the *x* sampling periods resulting in a reduction of adult returns between ***xxx*** and ***x,xxx***. We suggest that modifications to the habitat due to human activities in the Deadwater Reach has resulted in favorable conditions for northern pikeminnow, including improving conditions for predation upon juvenile Chinook salmon, which has resulted in a consequential impact to ESA-listed Chinook salmon populations in the Upper Salmon River MPG.

@@

Although not formally assessed here, another potential source of predation upon juvenile salmon in the Deadwater Slough is from avian predators including great blue herons *Ardea herodias* and bald eagles *Haliaeetus leucocephalus*. The Deadwater Slough is recognized as an important bird watching and nesting area due to the riparian and backwater habitats created by the feature [(Deadwater Slough - Audubon Important Bird Areas)](https://www.audubon.org/important-bird-areas/deadwater-slough#). During the initial sampling event in 2019, a two-person crew equipped with a mobile passive integrated transponder (PIT) tag antenna walked the entire reach and surrounding and upstream areas scanning for PIT tags. During the informal survey, we did identify nine PIT tags on surfaces surrounding Deadwater Slough, many of which were near nests or in an upstream anastomizing reach where herons and eagles are prevalent. Investigating their tag histories in [PTAGIS](https://www.ptagis.org/), these tags were implanted into a combination of juvenile Chinook Salmon (3), Sockeye Salmon (3), and steelhead (3) suggesting avian predation does occur at and adjacent to the Deadwater Reach, although to an unknown degree. ***Follow up with a brief lit review on avian predation on juvenile Chinook salmon elsewhere***

>>>>

```{r load-libraries}
library(tidyverse)
library(magrittr)
library(lubridate)
library(tidytable)
library(janitor)
library(purrr)
library(here)
library(FSA)

@@

The Upper Salmon River major population group (MPG) supports eight independent, extant spring/summer Chinook Salmon *Oncorhynchus tshawytscha* populations including Salmon River (above Redfish Lake), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. At least five of these eight populations must meet criteria set forth by @McElhany2000 and @ICTRT2007 for the MPG to be considered viable and for the recovery of the Snake River Evolutionary Significant Unit (ESU). Populations within the ESU have substantial cultural value, support downriver mainstem Snake and Columbia River commercial and subsistence fisheries, and support local fisheries and economies in years with sufficient abundance. All populations within the Upper Salmon River MPG have become depleted in recent decades. Declines in survival of juvenile Chinook Salmon have been attributed to the removal of beavers from the landscape (fur trade), mining activities, river simplification, water withdrawals, logging activities, urbanization, avian predation, proliferation of non-native species (e.g., non-native coastal rainbow trout *O. mykiss irideus* and brook trout *Salvelinus fontinalis*), warming streams and rivers, and modifications to downriver migration corridors (e.g., from hydropower projects). Moreover, the abundance of returning adults are further impacted by ocean and downriver harvests, poor ocean conditions, and changes to the spawning migration corridor. Each of these factors have contributed, to varying and unknown extents, to reduced adult escapement, the primary metric used to assess population viability. In response to the decline in Chinook Salmon abundance from the myriad human activities and associated habitat degradation, action agencies have attempted to improve juvenile survival and adult spawning conditions by investing in the rehabilitation of tributary ecosystems.

@@

```{r schnabel-estimator, eval = F}
# using Schnabel model
N_mods = cmr_df %>%
  filter(species == "Northern Pikeminnow") %>%
  # filter out spring 2021 (no recapture)
  filter(event_name != "Spring_2021") %>%
  group_by(event_name, species, date, fish_status, capture_type) %>%
  summarise(n_fish = n(),
            .groups = "drop") %>%
  group_by(event_name, species, date) %>%
  summarise(n = sum(n_fish),
            m = sum(n_fish[capture_type == "Recapture"]),
            R = sum(n_fish[fish_status == "Live"]),
            dead_fish = sum(n_fish[fish_status == "Dead"]),
            .groups = "drop") %>%
  group_by(event_name, species) %>%
  nest() %>%
  mutate(mod = map(data,
                   .f = function(x) {
                     with(x,
                          mrClosed(n = n,
                                   m = m,
                                   R = R,
                                   method = "Schnabel",
                                   chapman.mod = TRUE))
                   }),
         N = map_dbl(mod,
                     .f = summary),
         CI = map(mod,
                  .f = confint))

@@

We estimated the population size of Northern Pikeminnow in the Deadwater Slough to be greater than ***xx,xxx*** during the fall emigration period for DSR Chinook Salmon. That estimate translates to a density of ***xxx*** Northern Pikeminnow per 100 m or ***xxx*** per 100 m^2^ which is ***similar/more/less*** than estimates from elsewhere in the Columbia River ***(citation)*** where substantial Northern Pikeminnow predation impacts on salmonids have led to bounty programs aimed at reducing Northern Pikeminno abundance. The population size of Northern Pikeminnow was not directly estimated during the spring NRR Chinook salmon emigration period however, the relative abundance measured at CPUE was comparable to the fall sampling periods ***(update statement later)***. The population of Northern Pikeminnow in Deadwater Slough was estimated to consume between ***xx,xxx*** and ***xx,xxx*** juvenile Chinook Salmon during the ***x*** sampling periods and result in an estimated reduction of returning adults between ***xxx*** and ***x,xxx***. We suggest that the habitat modifications that created the Deadwater Slough reach have resulted in favorable conditions for Northern Pikeminnow, including improved conditions for predation upon juvenile Chinook Salmon ***(add detail here)***. Therefore, predation by Northern Pikeminnow in the Deadwater Slough likely has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon River MPG.

@@

Although not formally assessed in this study, avian predators including Great Blue Herons *Ardea herodias* and Bald Eagles *Haliaeetus leucocephalus* are another potential source of mortality for juvenile salmon in the Deadwater Slough. The Deadwater Slough is recognized as an important bird watching and nesting area due to the riparian and backwater habitats created by the feature [(Deadwater Slough - Audubon Important Bird Areas)](https://www.audubon.org/important-bird-areas/deadwater-slough#). Several piscivorous bird species have been documented using the Deadwater Slough that include the Common *Mergus merganser* and Hooded *Lophodytes cucullatus* mergansers, the Great Blue Heron, the Double-crested Cormorant *Phalacrocorax auritus*, and the Belted Kingfisher *Megaceryle alcyon* [@eBird2021]. During the intial sampling event in 2019, a two-person crew walked the entire reach and surrounding and upstream areas scanning for passive integrated transponder (PIT) tags. During that informal survey, nine PIT tags were recovered near active bird nests or in an upstream anastomizing reach where herons and eagles are prevalent, suggesting that mortality may have been a result of avian predation. The PIT tag histories in [PTAGIS](https://www.ptagis.org/) indicate these tags were implanted into a combination of juvenile Chinook Salmon (3), Sockeye Salmon (3), and steelhead (3). Avian predation contributes a major component of the total mortality for yearling Chinook Salmon in some locations in the lower Snake River and Columbia River, particularly at hydroelectric dams and within reservoirs [@Evans2012, @Evans2016]; however, we did not observe large colonies of piscivorous birds within the study area. Although there is documentation of individual Double-crested Cormorants [@eBird2021] at the Deadwater Slough, the site is not within their breeding range, rather, it is part of a migration corridor. Given the current avian species known to occupy Deadwater Slough, it is unlikely that avian predation on juvenile salmonids is comparable to elsewhere in the Columbia River basin with large piscivorous bird colonies. Nevertheless, we hypothesize that the reservoir-like conditions at the Deadwater Slough may increase the probability of avian predation on juvenile Chinook Salmon from the many piscivorous birds known to use the site. Future estimates of predation would benefit from consideration of the contribution of piscivorous avian predators.
